---
title: "STAT 407 - Principal Component Analysis Example"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center')
```

# Women's Heptathalon Data - 1988 Olympics

Example from Chapter 3, Section 10 of textbook

```{r pkgs}
library(tidyverse)
library(GGally)
heptath <- read_csv("heptathalonData.csv")
heptath$Medal[is.na(heptath$Medal)] <- "none"
head(heptath)
```

## Scores or PCAs? 

There are 7 events, and each competitor is assigned points for their performance in each event to get an overall score to win the medals. 

We want to perform PCA to see how closely the scores assigned match the results from the points. 

## Plotting the Data

```{r pairs}
ggpairs(heptath, 3:10) + ggtitle("Correlations between Heptathalon Events")
```


```{r medals}
ggplot(data = heptath) + 
  geom_point(aes(x = javelin, y = score, color = Medal), size = 3) + 
  scale_color_manual(values = c("#965A38","#D9A441", "grey20","#A8A8A8"), labels = c("bronze", "gold", "none", "silver")) + 
  theme_classic() + theme(legend.position = 'none') + ggtitle("Overall score by javelin distance")
```

## Performing PCA

```{r pca, echo = T, fig.align='center'}
heptath_pca <- prcomp(heptath[,3:9], scale = TRUE) 
print(heptath_pca)
```

Percent of variation in data explained by the PCs

```{r pcasd, echo = TRUE}
perc_var <- heptath_pca$sdev^2 / sum(heptath_pca$sdev^2)
round(perc_var * 100 , 2)
```

## PCA Plots 

```{r skree, echo=T, fig.align='center'}
perc_var <- heptath_pca$sdev^2 / sum(heptath_pca$sdev^2)
qplot(x = 1:7, y = perc_var, geom = 'line') + theme_classic() + 
  labs(x = "Principal Component", y = "Proportion of variation accounted for", title = "Scree plot")
```

## Where are the medal winners? 

```{r pcaplot, echo = FALSE}
heptath$PC1 <- heptath_pca$x[,1]
heptath$PC2 <- heptath_pca$x[,2]

ggplot(data = heptath) + 
  geom_point(aes(x = PC1, y = PC2)) + 
  theme_classic() + 
  coord_fixed() + 
  labs(x = "PC1", y = "PC2", title = "Data projected into the PCA space")
```

```{r pcaplot2, echo = FALSE}
ggplot(data = heptath) + 
  geom_point(aes(x = PC1, y = PC2, color = Medal), size = 3) +
  scale_color_manual(values = c("#965A38","#D9A441", "grey20","#A8A8A8"), labels = c("bronze", "gold", "none", "silver")) + 
  theme_classic() + 
  coord_fixed() + theme(legend.position = 'none') +  
  labs(x = "PC1", y = "PC2", title = "Data projected into the PCA space") 
```

## How do PCs compare to Score? 

```{r corpca, echo=TRUE}
cor(heptath$score, heptath$PC1)
```

```{r corpca2}
ggplot(data = heptath) + 
  geom_point(aes(x = PC1, y = score, color = Medal), size = 3) +
  scale_color_manual(values = c("#965A38","#D9A441", "grey20","#A8A8A8"), labels = c("bronze", "gold", "none", "silver")) + 
  theme_classic() + theme(legend.position = 'none') +  
  labs(x = "PC1", y = "Olympic Score", title = "Olympic Score by PCA") 
```